A performance assessment of evolutionary algorithms in volunteer computing environments: the importance of entropy

### 2016-11-15 EvoPar2017/paper.Rnw
Incorporates excellent fixes by @unintendedbear
Also eliminates .tex to avoid confusion. All changes should go to the .Rnw file.

Using this approach for creating distributed experiments is called {\em
  volunteer}, {\em cycle-scavenging} or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the web \cite{david-seti:home}. Our interest
here, however, is to use it as a resource for evolutionary
computation, as our group has done for a long time \cite{jj-ppsn98}.

In this line of research that uses volunteer computing for
evolutionary algorithms, there are several pending issues. The first
and maybe most important is approaching volunteer computing as a
socio-technical system \cite{vespignani2009predicting} which integrates
user decisions and behavioral patterns in the system model; this
includes trying to optimize the number of users in a particular
experiment. The second line of research, although related to the
first, is more focused on the evolutionary algorithm itself and how
different design decisions will affect its performance.
% Is it ok that we say that the first line of research is the most important... but then we focus on the second?
We have approached the first issue in our previous work \cite{jj-ppsn98anonymous},
% I have justified to focus, in this case, on the second
but in this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all available resources for finding the solution of an evolutionary algorithm. This design will be done incrementally by
changing client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions and is able to find
the solution to the problem in a minimum time and with the least
number of contributions.

@@

Volunteer computing involves users deciding to run a program that acts
as a client or as a peer in an experiment
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home} or
high-throughput queuing system that such as HTCondor \cite{HTCondor}.
However
the dual facts of the introduction of JavaScript as a universal language for the
browser and the browser itself as an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in [citation hidden]. Systems based on the
JavaScript/browser combination
emphasize the ephemerality, ease of use and universality, while
systems such as HTCondor or BOINC might be more adequate for work that
require higher availability of volunteer resources, achieving this
with downloadable clients, in some cases like HTCondor needing
administration access to the resources.
%\cite{DBLP:conf/gecco/GuervosG15}.

@@

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used. 
The evolutionary algorithm that uses Trap as fitness function is
completely run in the
browser, but the server needs to know when the solution has been
found, so a subroutine that checks if the solution been found  
is also included in the server, as a configuration
option. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.
For every new individual sent by the clients to the cache, several
pieces of metadata are stored: a time stamp, the client IP, the chromosome and
fitness value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

All experiments are announced in the same way: we use social networks,
mainly Twitter, to announce the start of a new experiment. LinkedIn,
Facebook, and also private groups in Telegram and WhatsApp have been
used for announcements too. We also encourage friends and followers to
put an original post for announcing it, instead of just sharing, in
the way allowed by the platform, the original post. During the running
of the experiments, users are engaged, questions answered, and all
kind of explanations given if requested. In general, announcements
have been made in Spanish, but in some cases English was used
too. This makes for a certainly unreliable experimentation framework,
but it is a realistic one. The steps in the design of the evolutionary
algorithm will be explained next.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using a volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to nil. In fact, the NodIO system can be deployed to any
Platform or Infraestructure as a Service 
system as long as it can run node.js and can store logs on the same
place or in some other Log as a Service. Local storage can be used,
but it is not a requisite and, in fact, Heroku does not allow
persistent storage of files in its service. 
All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=6>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Log-log plot of ranked and normalized number of contributions (PUTs) per IP.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and thus having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, especially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. However, this was not known to the volunteer, so the fact that they were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize there was no contribution and do something
about it. These sets of runs are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if it found a new optimum, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently started. This batch of runs is labelled {\tt
  reboot}. This size of the cache was not really noticed by some slow clients or too fast runs, so that in the
last case, is labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage user engagement and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thing we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache implying a new run, this way overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that some clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which
needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} is done only when the cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

In fact, Table \ref{tab:correlation}, shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one. A high correlation indicating that the number of volunteers changes
slowly from one minute to the next; however, the {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with volunteers joining and leaving the experiment all the
time, maybe because experiments take less time and need less resources.

@@

The graph that plots the time to solution of every instance of the
experiment vs. entropy is shown in \ref{fig:entropy}. There is a clear
trend towards experiments with lower entropy taking more time and
experiments with higher entropy taking more time, which indeed can be
statistically fit to a linear model. This dependence on entropy is
much clearer than the assumed dependence on the number of users, which
is represented in the graph by dot color, and even the number of PUTs
or contributions, which is represented by dot size. You can see a
clearly blue and big dot, meaning more than 40 different IPs
participating and around 1500 PUTs in the high end of the graph,
showing that it took the longest to find the solution. On the other
hand, clearly red and small dots are on the lowest end of the graph,
indicating that it took a short time to find the solution as long as
the entropy was high.

@@

Our main intention in this paper was to configure an evolutionary
algorithm in a volunteer computing environment to improve its
efficiency and eventually find out on what this efficiency depends
on. The baseline configuration that we had used in other papers 
\cite{jj-ppsn98anonymous} included
{\em overlaps} or volunteers running islands that had actually started in
a previous experiment instance in the set. In previous papers, we were
interested in measuring the number of users or their behavior in bulk
and this overlap was not so important.
In this case our focus was on the evolutionary algorithm so we had to
make all volunteers in it started more or less at the same time and
from a random population. %Pablo: too long paragraph, added commas and sepparated into two
In fact, we managed to do so, except in some
byzantine cases, by making the clients restart when the cache size
collapses to 1 or to a size much smaller than the current; we called
this client configuration {\tt reboot} and {\tt cachecrash}.  This
configuration achieves a positive secondary effect: all clients
contribute to fill the cache, having thus an actual effect on the
chromosome pool and effectively making all clients present contribute
to the simulation.

>>>>

Using this approach for creating distributed experiments is called {\em
  volunteer}, {\em cycle-scavenging}, or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the web \cite{david-seti:home}. Our interest
here, however, is to use it as a resource for evolutionary
computation, as our group has done for a long time \cite{jj-ppsn98anonymous}.

In this line of research that uses volunteer computing for
evolutionary algorithms, there are several pending issues. The first
and maybe most important is approaching volunteer computing as a
socio-technical system \cite{vespignani2009predicting} which integrates
user decisions and behavioral patterns in the system model; this
includes trying to optimize the number of users in a particular
experiment. The second line of research, although related to the
first, is more focused on the evolutionary algorithm itself and how
different design decisions will affect its performance.
% Is it ok that we say that the first line of research is the most important... but then we focus on the second?
We have approached the first issue in our previous work \cite{jj-ppsn98anonymous},
% I have justified to focus, in this case, on the second
but in this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all available resources for finding the solution of an evolutionary algorithm. This design will be done incrementally by
changing client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions and is able to find
the solution to the problem in a minimum time, with the least
number of contributions.

@@

Volunteer computing involves users deciding to run a program that acts
as a client or as a peer in an experiment
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}, or a 
high-throughput queuing system such as HTCondor \cite{HTCondor}.
However,
the dual facts of the introduction of JavaScript as a universal language for the
browser and the browser itself as both an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15anonymous}. Systems based on the
JavaScript/browser combination
emphasize the ephemerality, ease of use, and universality, while
systems such as HTCondor or BOINC might be more adequate for work that % citation for BOINC (?)
require higher availability of volunteer resources, achieving this
with downloadable clients which in some cases, like HTCondor, need
administration access to the resources.

@@

NodIO needs a fitness function to work with, in this case the classical Trap 
function \cite{Ackley1987} has been used. 
The evolutionary algorithm that uses Trap as fitness function is
completely run in the
browser, but the server also includes, as a configuration
option, a subroutine that checks if the solution has been
found. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes. % Why? Do we have results on a single computer or on a cluster to compare?
For every new individual the clients send to the cache, several
pieces of metadata are stored: a time stamp, the client IP, the chromosome and
fitness value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

All experiments are announced in the same way: we use social networks,
mainly Twitter, to announce the start of a new experiment. LinkedIn,
Facebook, and also private groups in Telegram and WhatsApp have been
used for announcements too. We also encourage friends and followers to
put their own posts for announcing it, instead of just sharing, in
the way allowed by the platform, our original ones. For instance, when using Twitter, to include the URL of the experiment inside a tweet of their own, instead of just sharing it via a re-tweet (RT). During the running
of the experiments, users are engaged, questions answered, and all
kind of explanations given if requested. In general, announcements
have been made in Spanish, but in some cases English was used
too. This makes for a certainly unreliable experimentation framework,
but it is a realistic one. The steps in the design of the evolutionary
algorithm will be explained next.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using a volunteer computing system is to cut costs, %actually, we talk about the system being efficient, and also the algorithm itself, but I think we don't talk about the costs...
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to nil. In fact, the NodIO system can be deployed to any
Platform or Infrastructure as a Service 
system as long as it can run \texttt{node.js} and can store logs on the same
place or in some other Log as a Service. Local storage can be used,
but it is not a requisite and, in fact, Heroku %first time Heroku is mentioned...
 does not allow
persistent storage of files in its service. 
All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=6>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Log-log plot of ranked and normalized number of contributions (PUTs) per IP.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request in Figure \ref{fig:system}) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and thus having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, especially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph which shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. However, this was not known to the volunteer, so the fact that they were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize there was no contribution and do something
about it. These sets of runs are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if it found a new optimum, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently started. This batch of runs is labelled {\tt
  reboot}. This size of the cache was not really noticed by some slow clients or too fast runs, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thing we will have to check is whether we have effectively eliminated those
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established the
minimum at 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. % Why?
 Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, which shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which
needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} only does so when cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

In fact, Table \ref{tab:correlation}, which shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one. A high correlation indicating that the number of volunteers changes
slowly from one minute to the next; however, the {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with volunteers joining and leaving the experiment all the
time, maybe because experiments take less time and need less resources.

@@

The graph that plots the time to solution of every instance of the
experiment vs. entropy is shown in \ref{fig:entropy}. There is a clear
trend towards experiments with lower entropy taking more time and
experiments with higher entropy taking more time, which indeed can be
statistically fit to a linear model. This dependence on entropy is
much clearer than the assumed dependence on the number of users, which
is represented in the graph by dot color, and even the number of PUTs
or contributions, which is represented by dot size. You can see a
clearly blue and big dot, meaning more than 40 different IPs
participating and around 1500 PUTs in the high end of the graph,
showing that it took the longest to find the solution. On the other
hand, clearly red and small dots are on the lowest end of the graph,
indicating that it took a short time to find the solution as long as
the entropy is high.

@@

Our main intention in this paper was to configure an evolutionary
algorithm in a volunteer computing environment to improve its
efficiency and eventually find out what this efficiency depends
on. The baseline configuration that we had used in other papers  included
{\em overlaps} or volunteers running islands that had actually started in
a previous experiment instance in the set. In previous papers, we were
interested in measuring the number of users or their behavior in bulk
and this overlap was not so important, but
in this case our focus was on the evolutionary algorithm so we had to
make all volunteers in it started more or less at the same time and
from a random population. %Pablo: too long paragraph, added commas and sepparated into two
In fact, we managed to do so, except in some
byzantine cases, by making the clients restart when the cache size
collapses to 1 or to a size much smaller than the current; we called
this client configuration {\tt reboot} and {\tt cachecrash}.  This
configuration achieves a positive secondary effect: all clients
contribute to fill the cache, having thus an actual effect on the
chromosome pool and effectively making all clients present contribute
to the simulation.
